|

### 2025-05-24 02-methods.Rmd
add assets + edits methods

## Background Measures

As an initial step in our analysis, and to better understand students' awareness of and attitudes toward AI, we developed a brief four-question "AI in Education Attitudes Scale." This scale was adapted from the broader AI Attitudes Scale proposed by Schepman and Rodway [@schepman2020initial]. These questions included: Do you agree or disagree that society will benefit from a future of AI? Do you agree or disagree that AI is dangerous? Do you agree or disagree that AI will foster students’ learning in the future? Do you agree or disagree that AI is often misused by students?

Finally, to consider participants' academic performance and skills, participants were also asked about their school grades and how often they complete their homework assignments on time and what factors affect their ability to do so. Self-reported experience using ChatGPT (the main LLM available at the time of the study) was also asked, including a question on their estimation of how many of their peers use ChatGPT for their homework.

Section \@ref(sec:questionnaire) presents the complete questionnaire.

>>>>

## Student Characteristics

To examine student interactions with AI and their potential impact on learning and critical thinking, we collected a range of self-reported and behavioral data. The full questionnaire is provided in the Appendix (Section \@ref(sec:questionnaire)).

## AI Attitudes and Usage

To better understand students' general attitudes toward AI, we selected four questions from Schepman and Rodway's AI Attitudes Scale [@schepman2020initial]. These questions explored students; beliefs about AI's societal benefits, potential dangers, capacity to support learning, and likelihood of misuse by students (Appendix, Section \@ref(sec:questionnaire)). We also asked students about their direct experience using ChatGPT for homework --- the most common chatbot at the time of the study --- and their beliefs of how many of their peers use ChatGPT for their homework, thereby capturing both individual behaviour and the perceived social norm around AI usage. These measures help establish baseline orientations that may moderate how students engage with and evaluate AI explanations (related to RQ-1) and how they perceive different AI interaction styles (related to RQ-3).

## AI Academic Habits and Skills

Academic skills or habits may moderate the effect of AI interactions on problem solving activities. Accordingly, to acconut for participants' heterogeneity in academic skills, we asked students to report their average school grades within five categories. We also asked students about their academic habits or the challenges they face at school, such as how often they complete their homework assignments on time and what factors affect their ability to do so.

## AI Interactions

To quantify engagement, we analyzed students' interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity. While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or depth of reasoning, limiting its direct relevance to RQ-1 and RQ-2. Therefore, we also asked students about how helpfuly they found the AI interactions and how confident they were in their answers to selected tasks. These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ-1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ-2).

### 2025-05-24 03-results.Rmd
add assets + edits methods

Table \@ref(tab:summary-statistics) illustrates the distribution of student characteristics across treatment groups alongside the results of separate Fisher's Exact Tests on the association with treatment assignment. The sample characteristics were generally balanced across treatments. Only one variable out of ten -- i.e., self-reported difficulties in homework -- was statistically associated (p = 0.05) with the treatment assignment to Socratic/Non-Socratic conditions, while no significant associations were found for the AI explanation assignment. Thus, only one out of twenty Fisher's exact tests showed a significant result, indicating a good balance across treatments.

@@

Conversely, exposure to the AI step-by-step reasoning seemed associated with students' perception of the accuracy of the "crowd" guess --- the average guess from 600 participants reported in the original study [@steiner2015turns].^[Notably, this guess was $596, exceeding the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213).] All groups received the same human and AI estimates, but while 64% of students in the control group rated the human estimate as "poor" or "very poor," only 45% of students exposed to AI reasoning did so. Although this difference was insignificant (Fisher's exact test, p = 0.15),  regression analysis, controlling for student's location, gender, their initial estimate of the value of coins, and experience with ChatGPT, showed a significant effect (p < 0.05) of AI exposure on students' perceptions of the human estimate, as shown in Figure \@ref(fig:perceived-accuracy-coins). This finding underscores the complex relationship between AI and learning. It suggests that students exposed to the AI's step-by-step reasoning may have identified minor errors, thus increasing their expectations of human accuracy, and those who viewed AI as a "black box" tended to underrate human estimates.

>>>>

Table \@ref(tab:summary-statistics) illustrates the distribution of student characteristics across treatment groups alongside the results of separate Fisher's Exact Tests on the association with treatment assignment. The sample characteristics were generally balanced across treatment groups. Only one variable out of ten -- i.e., self-reported difficulties in homework -- was statistically associated (p = 0.05) with the Socratic/Non-Socratic assignment, while no significant associations were found for the AI-reasoning assignment. Thus, only one out of twenty Fisher's exact tests showed a significant result, indicating a good balance across treatments.

@@

Conversely, exposure to the AI step-by-step reasoning seemed associated with students' perception of the accuracy of the "human" guess --- the average guess from 600 participants reported in the original study [@steiner2015turns].^[Notably, this guess was $596, exceeding the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213).] While all groups received the same human and AI estimates, 64% of students in the control rated the human estimate as "poor" or "very poor" against 45% of students exposed to AI reasoning. Although this difference was insignificant (Fisher's exact test, p = 0.15),  regression analysis, controlling for fixed differences across schools and across individuals, such as their gender, the initial guess of the value of coins, and experience with ChatGPT, showed a significant effect (p < 0.05) of the assignment to AI reasoning on students' perceptions of the human guess, as shown in Figure \@ref(fig:perceived-accuracy-coins). This finding underscores the complex relationship between AI and learning. It suggests that students exposed to the AI's step-by-step reasoning may have identified minor errors, thus increasing their expectations of human accuracy, and those who viewed AI as a "black box" tended to underrate human estimates.
